Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants

### 2022-06-25 ms/ms.tex
r1

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations and generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain covariance between stomatal anatomical traits. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this synthesizing data on stomatal density and length from amphistomatous leaves of 638 terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 236 phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems (\protect\hyperlink{ref-mackenzie_proving_1999}{Mackenzie 1999}; \protect\hyperlink{ref-maritan_optimal_2000}{Maritan et al. 2000}). Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a ``disposition for covariation'' (\protect\hyperlink{ref-armbruster_integrated_2014}{Armbruster et al. 2014}), meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution (reviewed in \protect\hyperlink{ref-pelabon_evolution_2014}{Pélabon et al. 2014}). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection (\protect\hyperlink{ref-lande_quantitative_1979}{Lande 1979}). However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection (\protect\hyperlink{ref-hansen_is_2003}{Hansen 2003}). For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg's rule, \protect\hyperlink{ref-berg_general_1959}{Berg 1959}, \protect\hyperlink{ref-berg_ecological_1960}{1960}; \protect\hyperlink{ref-conner_raissa_2014}{Conner and Lande 2014}; \protect\hyperlink{ref-armbruster_covariance_1999}{Armbruster et al. 1999}).

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of land plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of year of evolutionary history. A heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may potentially differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological importance. The next sections provides background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO\(_2\) gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants (recently reviewed \protect\hyperlink{ref-clark_origin_2022}{Clark et al. 2022}). Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status (\protect\hyperlink{ref-hetherington_role_2003}{Hetherington and Woodward 2003}; \protect\hyperlink{ref-lawson_guard_2020}{Lawson and Matthews 2020}). On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO\(_2\) by altering stomatal anatomy in new leaves (\protect\hyperlink{ref-casson_influence_2008}{Casson and Gray 2008}). Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change (\protect\hyperlink{ref-haworth_co-ordination_2013}{Haworth, Elliott-Kingston, and McElwain 2013}). Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are likely both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

@@

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous, suggesting that the benefits of amphistomy. Amphistomy should increase photosynthesis most under saturating-light conditions where CO\(_2\) supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}), especially in herbs (\protect\hyperlink{ref-muir_light_2018}{Muir 2018}). However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants (\protect\hyperlink{ref-muir_is_2019}{Muir 2019}), suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in \emph{Populus} (\protect\hyperlink{ref-mckown_association_2014}{McKown et al. 2014}, \protect\hyperlink{ref-mckown_role_2019}{2019}; \protect\hyperlink{ref-fetter_growthdefense_2021}{Fetter, Nelson, and Keller 2021}). Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits (\protect\hyperlink{ref-buckley_how_2015}{Buckley et al. 2015}). Other hypotheses about the adaptive significance of stomatal ratio are discussed in Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) and Drake et al. (\protect\hyperlink{ref-drake_two_2019}{2019}).

@@

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as ``inverse size-density scaling'' and ``bimodal stomatal ratio'' (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata (\protect\hyperlink{ref-weiss_untersuchungen_1865}{Weiss 1865}; \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO\(_2\) and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO\(_2\) for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO\(_2\). Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO\(_2\) concentration gradient around one stomate merges with that of its neighbor (\protect\hyperlink{ref-zeiger_stomatal_1987}{Zeiger, Farquhar, and Cowan 1987}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}), incur metabolic costs (\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}), and/or risk hydraulic failure (\protect\hyperlink{ref-henry_stomatal_2019}{Henry et al. 2019}). The distribution of stomata and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}) further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal \ref{fig:concepts}b. Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Amphistomy occurs most often in herbaceous plants from open, high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}, \protect\hyperlink{ref-muir_light_2018}{2018}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}). Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (\emph{cf.} Peat and Fitter (\protect\hyperlink{ref-peat_comparative_1994}{1994}) Fig 1).

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in land plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Could packing limits and developmental integration explain? We limit our focus in this study to interspecific variation in mean trait values and do not study understand intraspecifc variation here.

Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increase as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}).

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}), the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}). For example, larger genomes increase meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration of stomatal anatomy (\protect\hyperlink{ref-beaulieu_genome_2008}{Beaulieu et al. 2008}; \protect\hyperlink{ref-simonin_genome_2018}{Simonin and Roddy 2018}; \protect\hyperlink{ref-roddy_scaling_2020}{Roddy et al. 2020}). Developmental integration in this case would not hinder adaptive evolution when the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}), then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints like contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection (\protect\hyperlink{ref-barrett_sexual_2013}{Barrett and Hough 2013}). We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We integrated and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).

>>>>

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain trait covariance. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this by synthesizing data on stomatal density and length from amphistomatous leaves of 638 terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 236 phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems (\protect\hyperlink{ref-mackenzie_proving_1999}{Mackenzie 1999}; \protect\hyperlink{ref-maritan_optimal_2000}{Maritan et al. 2000}). Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a ``disposition for covariation'' (\protect\hyperlink{ref-armbruster_integrated_2014}{Armbruster et al. 2014}), meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution (reviewed in \protect\hyperlink{ref-pelabon_evolution_2014}{Pélabon et al. 2014}). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection (\protect\hyperlink{ref-lande_quantitative_1979}{Lande 1979}). However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection (\protect\hyperlink{ref-hansen_is_2003}{Hansen 2003}). For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg's rule, \protect\hyperlink{ref-berg_general_1959}{Berg 1959}, \protect\hyperlink{ref-berg_ecological_1960}{1960}; \protect\hyperlink{ref-conner_raissa_2014}{Conner and Lande 2014}; \protect\hyperlink{ref-armbruster_covariance_1999}{Armbruster et al. 1999}).

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO\(_2\) gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants (recently reviewed in \protect\hyperlink{ref-clark_origin_2022}{Clark et al. 2022}). Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status (\protect\hyperlink{ref-hetherington_role_2003}{Hetherington and Woodward 2003}; \protect\hyperlink{ref-lawson_guard_2020}{Lawson and Matthews 2020}). On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO\(_2\) by altering stomatal anatomy in new leaves (\protect\hyperlink{ref-casson_influence_2008}{Casson and Gray 2008}). Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change (\protect\hyperlink{ref-haworth_co-ordination_2013}{Haworth, Elliott-Kingston, and McElwain 2013}). Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

@@

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous. Amphistomy should increase photosynthesis most under saturating-light conditions where CO\(_2\) supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}), especially in herbs (\protect\hyperlink{ref-muir_light_2018}{Muir 2018}). However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants (\protect\hyperlink{ref-muir_is_2019}{Muir 2019}), suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in \emph{Populus} (\protect\hyperlink{ref-mckown_association_2014}{McKown et al. 2014}, \protect\hyperlink{ref-mckown_role_2019}{2019}; \protect\hyperlink{ref-fetter_growthdefense_2021}{Fetter, Nelson, and Keller 2021}). Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits (\protect\hyperlink{ref-buckley_how_2015}{Buckley et al. 2015}). Other hypotheses about the adaptive significance of stomatal ratio are discussed in Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) and Drake et al. (\protect\hyperlink{ref-drake_two_2019}{2019}).

@@

Two major features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as ``inverse size-density scaling'' and ``bimodal stomatal ratio'' (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata (\protect\hyperlink{ref-weiss_untersuchungen_1865}{Weiss 1865}; \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO\(_2\) and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO\(_2\) for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO\(_2\). Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO\(_2\) concentration gradient around one stomate merges with that of its neighbor (\protect\hyperlink{ref-zeiger_stomatal_1987}{Zeiger, Farquhar, and Cowan 1987}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}), incur metabolic costs (\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}), and/or risk hydraulic failure (\protect\hyperlink{ref-henry_stomatal_2019}{Henry et al. 2019}). The distribution of stomatal size and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}) further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal (Fig. \ref{fig:concepts}b). Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Amphistomy occurs most often in herbaceous plants from open, high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}, \protect\hyperlink{ref-muir_light_2018}{2018}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}). Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (\emph{cf.} \protect\hyperlink{ref-peat_comparative_1994}{Peat and Fitter 1994, fig. 1}).

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in flowering plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecifc variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}) for the species' mean, not an individual leaf.

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}), the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}). For example, larger genomes increase meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy (\protect\hyperlink{ref-beaulieu_genome_2008}{Beaulieu et al. 2008}; \protect\hyperlink{ref-simonin_genome_2018}{Simonin and Roddy 2018}; \protect\hyperlink{ref-roddy_scaling_2020}{Roddy et al. 2020}). Developmental integration in this case would not necessarily hinder adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}), then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and/or developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints likely contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection (\protect\hyperlink{ref-barrett_sexual_2013}{Barrett and Hough 2013}). We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).
